---
title: "Plot data (initial analysis) - Project 3"
output: html_document
author: Saibal Bhattacharya
---

**Prework: Load the CSV file and libraries.**  
```{r}
library(gridExtra)
library(tidyverse)
library(data.table)
library(hrbrthemes)
library(lubridate)
library(chron)
library(data.table)
library(ggplot2)
library(repr)
library(dplyr)
library(caret)

# Set the initial plot area dimensions
options(repr.plot.width=4, repr.plot.height=4) 

df1 <- read.csv("PROJECT3_DATA.csv")

```

```{r}
# --------------- Initial data analysis and cleanup ----------------------------

df1 <- read.csv("PROJECT3_DATA.csv")
df4 <- df1 %>% filter(attemptresult != "SYSLOGOUT")
df5 <- setDT(df4)
df6 <- df5 %>% arrange(maxsessionuid, attemptdate)
df6b <- unique( df6[ , 1:7 ] )
df7 <- df6b[, order := seq(1:.N), by=maxsessionuid]
df7_login <- df7[attemptresult == "LOGIN"]
df7_logout <- df7[attemptresult == "LOGOUT" | attemptresult == "TIMEOUT" | attemptresult == "RESTART"]
df7_logout[, order := order - 1]
df8 <- df7_login[, .(maxsessionuid, order, type, userid, defsite, groupname, attemptdate)][df7_logout[, .(maxsessionuid, order, type, userid, defsite, groupname, attemptdate)], 
                                                                                           on=c("maxsessionuid", "order", "type", "userid", "defsite", "groupname"), nomatch=NA]
df9 <- setnames(df8, c("attemptdate", "i.attemptdate"), c("LOGIN_time", "Out_time"))
df10 <- df9[!is.na(df9$LOGIN_time), ]
df11 <- df10 %>% mutate(my_LOGIN_time = mdy_hm(df10$LOGIN_time), my_Out_time = mdy_hm(df10$Out_time))
df11$sessiontime <- round((df11$my_Out_time - df11$my_LOGIN_time)/60, 2)
df12 <- df11 %>% select(-LOGIN_time, -Out_time)
df12b <- df12[as.numeric(sessiontime) != 0]
df13 <- mutate_at(df12b, vars('type', 'defsite', 'groupname'), as.factor)
df13$login_hr <- hour(as.POSIXlt(df13$my_LOGIN_time))
df13$login_day <- day(as.POSIXlt(df13$my_LOGIN_time))
df13$login_wday <- weekdays(as.POSIXlt(df13$my_LOGIN_time))
df13$weekend = chron::is.weekend(df13$my_LOGIN_time)
df13$defsite_grp[df13$defsite == "TX"] <- "TX_all"
df13$defsite_grp[df13$defsite == "TX_PIPE"] <- "TX_all"
df13$defsite_grp[df13$defsite == "LA"] <- "LA_all"
df13$defsite_grp[df13$defsite == "LA_PIPE"] <- "LA_all"
df13$defsite_grp[df13$defsite == "OK_PIPE"] <- "OK_all"
df13b <- subset(df13, defsite_grp != "OK_all")
outlier <- function(x) {
  ifelse(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x), 
         TRUE, 
         FALSE)
}
df13c <- setDT(df13b)[, outlier_ident := outlier(sessiontime), by = 'defsite_grp']
df13_no_out <- df13c[outlier_ident == FALSE]

df15_no_out <- setDF(df13_no_out)
df15_no_out$weekend_num <- as.numeric(df15_no_out$weekend)
df15_no_out$sessiontime_num <- as.numeric(df15_no_out$sessiontime)
df15_no_out <- mutate_at(df15_no_out, vars('defsite_grp'), as.factor)

df15_no_out$groupname_bin[df15_no_out$groupname %in% c('IT SECURITY', 'IT ADMINISTRATOR', 'IT ANALYST')] <- 'IT'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('CORP PIPELINE TECH SUPPORT', 'PIPELINE FIELD MGMT','PIPELINE TECHNICIAN')] <- 'PIPELINE'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('CONTRACTOR PLANNER', 'PLANNER' , 'SCHEDULER')] <- 'PLANNING'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('CONTRACTOR', 'ENGINEER', 'QUICK REPORTING USER')] <- 'OPERATIONS'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('MASTER DATA', 'MAINTENANCE TECHNICIAN', 'CONTRACTOR TECHNICIAN', 'MAINTENANCE SUPERVISOR', 'CONTRACTOR SUPERVISOR')] <- 'MAINTENANCE'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('ACCOUNTING', 'PURCHASING MANAGER', 'SUPPLY CHAIN', 'STOREROOM MANAGER', 'STOREROOM CLERK', 'SENIOR STOREROOM CLERK')] <- 'SUPPLY CHAIN'

df15_no_out <- mutate_at(df15_no_out, vars('groupname_bin'), as.factor)

df15_no_out$gname_bin_num <- ifelse(df15_no_out$groupname_bin == "IT", 1,
                                    ifelse(df15_no_out$groupname_bin == "MAINTENANCE", 2,
                                           ifelse(df15_no_out$groupname_bin == "OPERATIONS", 3,
                                                  ifelse(df15_no_out$groupname_bin == "PIPELINE", 4,
                                                         ifelse(df15_no_out$groupname_bin == "PLANNING", 5,
                                                                ifelse(df15_no_out$groupname_bin == "SUPPLY CHAIN", 6,
                                                                       NA ))))))

df16_no_out <- df15_no_out

```

```{r}
# --------------------- Plot data --------------------------------------

#Plot all pairwise combinations of features with defsite_grp in colors
single_plot = function(df, colx, coly){
  ggplot(df, aes_string(colx,coly)) +
    geom_point(aes(color = factor(defsite_grp)), alpha = 0.4)
}

#Plot - login_hr vs. sessiontime_num 
plot_df16_no_out = function(df){
  options(repr.plot.width=8, repr.plot.height=5)
  grid.arrange(
    single_plot(df, 'sessiontime_num', 'login_hr'),
    nrow = 1)
}
plot_df16_no_out(df16_no_out)

```
<br>
**Observations:** 
1. Most logins occur between 5 AM and 6 PM with session times varying between 1 to < 100 mins.  
2. Users from TX_all dominate longer session times.  
3. Fewer login sessions occur during weekends, and are normally shorter.  

```{r}
#Plot - weekend_num vs. sessiontime_num 
plot_df16_no_out = function(df){
  options(repr.plot.width=8, repr.plot.height=5)
  grid.arrange(
    single_plot(df, 'sessiontime_num', 'weekend_num'),
    nrow = 1)
}
plot_df16_no_out(df16_no_out)

```
<br>
**Observations:**    
1. Whether it be weekday or weekend, longer login times come from TX_all

```{r}
#Plot - login_wday vs. sessiontime_num 
plot_df16_no_out = function(df){
  options(repr.plot.width=8, repr.plot.height=5)
  grid.arrange(
    single_plot(df, 'sessiontime_num', 'login_wday'),
    nrow = 1)
}
plot_df16_no_out(df16_no_out)

```
<br>
**Observations:**  
1. Longer session times are fewer on Saturdays and Sundays.

```{r}
#Plot - groupname_bin vs. sessiontime_num 
plot_df16_no_out = function(df){
  options(repr.plot.width=8, repr.plot.height=5)
  grid.arrange(
    single_plot(df, 'sessiontime_num', 'groupname_bin'),
    nrow = 1)
}
plot_df16_no_out(df16_no_out)

```
<br>
**Observation:**  
1. Shorter and fewer login session times come from folks in pipeline.  
2. In comparison to others (except pipeline), IT has folks show fewer longer session times.

```{r}
#Plot - groupname_bin vs. login_hr 
plot_df16_no_out = function(df){
  options(repr.plot.width=8, repr.plot.height=5)
  grid.arrange(
    single_plot(df, 'login_hr', 'groupname_bin'),
    #single_plot(df, 'weekend_num', 'groupname_bin'),
    nrow = 1)
}
plot_df16_no_out(df16_no_out)

```
<br>
**Observation:**  
1. Pipeline folks normally login between 6 AM and 3 PM.  
2. IT folks normally login between 5 AM and 5 PM.
3. Folks from other departments login anytime between zero and 24 hrs.

```{r}
#Plot - groupname_bin vs. weekend_num 
plot_df16_no_out = function(df){
  options(repr.plot.width=8, repr.plot.height=5)
  grid.arrange(
    single_plot(df, 'weekend_num', 'groupname_bin'),
    nrow = 1)
}
plot_df16_no_out(df16_no_out)

```
<br>
**Observation:**  
1. Folks from all groupname_bin login both during weekend and weekday.










