---
title: "NULL Hypothesis: Sessiontime_num - Question 2"
output: html_document
author: Saibal Bhattacharya
---

```{r }
library(tidyverse)
library(data.table)
library(hrbrthemes)
library(lubridate)
library(chron)
library(data.table)
library(ggplot2)
library(repr)
library(dplyr)
library(caret)
install.packages('effsize')
install.packages('ggridges')
library(effsize)
library(ggridges)
# Set the initial plot area dimensions
options(repr.plot.width=4, repr.plot.height=4) 

df1 <- read.csv("PROJECT3_DATA.csv")
```


```{r }
df4 <- df1 %>% filter(attemptresult != "SYSLOGOUT")
df5 <- setDT(df4)
df6 <- df5 %>% arrange(maxsessionuid, attemptdate)
df6b <- unique( df6[ , 1:7 ] )
df7 <- df6b[, order := seq(1:.N), by=maxsessionuid]
df7_login <- df7[attemptresult == "LOGIN"]
df7_logout <- df7[attemptresult == "LOGOUT" | attemptresult == "TIMEOUT" | attemptresult == "RESTART"]
df7_logout[, order := order - 1]
df8 <- df7_login[, .(maxsessionuid, order, type, userid, defsite, groupname, attemptdate)][df7_logout[, .(maxsessionuid, order, type, userid, defsite, groupname, attemptdate)], 
                                                                                           on=c("maxsessionuid", "order", "type", "userid", "defsite", "groupname"), nomatch=NA]
df9 <- setnames(df8, c("attemptdate", "i.attemptdate"), c("LOGIN_time", "Out_time"))
df10 <- df9[!is.na(df9$LOGIN_time), ]
df11 <- df10 %>% mutate(my_LOGIN_time = mdy_hm(df10$LOGIN_time), my_Out_time = mdy_hm(df10$Out_time))
df11$sessiontime <- round((df11$my_Out_time - df11$my_LOGIN_time)/60, 2)
df12 <- df11 %>% select(-LOGIN_time, -Out_time)
df12b <- df12[as.numeric(sessiontime) != 0]
df13 <- mutate_at(df12b, vars('type', 'defsite', 'groupname'), as.factor)
df13$login_hr <- hour(as.POSIXlt(df13$my_LOGIN_time))
df13$login_day <- day(as.POSIXlt(df13$my_LOGIN_time))
df13$login_wday <- weekdays(as.POSIXlt(df13$my_LOGIN_time))
df13$weekend = chron::is.weekend(df13$my_LOGIN_time)
df13$defsite_grp[df13$defsite == "TX"] <- "TX_all"
df13$defsite_grp[df13$defsite == "TX_PIPE"] <- "TX_all"
df13$defsite_grp[df13$defsite == "LA"] <- "LA_all"
df13$defsite_grp[df13$defsite == "LA_PIPE"] <- "LA_all"
df13$defsite_grp[df13$defsite == "OK_PIPE"] <- "OK_all"
df13b <- subset(df13, defsite_grp != "OK_all")
outlier <- function(x) {
  ifelse(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x), 
         TRUE, 
         FALSE)
}
df13c <- setDT(df13b)[, outlier_ident := outlier(sessiontime), by = 'defsite_grp']
df13_no_out <- df13c[outlier_ident == FALSE]

df15_no_out <- setDF(df13_no_out)
df15_no_out$weekend_num <- as.numeric(df15_no_out$weekend)
df15_no_out$sessiontime_num <- as.numeric(df15_no_out$sessiontime)
df15_no_out <- mutate_at(df15_no_out, vars('defsite_grp'), as.factor)

df15_no_out$groupname_bin[df15_no_out$groupname %in% c('IT SECURITY', 'IT ADMINISTRATOR', 'IT ANALYST')] <- 'IT'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('CORP PIPELINE TECH SUPPORT', 'PIPELINE FIELD MGMT','PIPELINE TECHNICIAN')] <- 'PIPELINE'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('CONTRACTOR PLANNER', 'PLANNER' , 'SCHEDULER')] <- 'PLANNING'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('CONTRACTOR', 'ENGINEER', 'QUICK REPORTING USER')] <- 'OPERATIONS'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('MASTER DATA', 'MAINTENANCE TECHNICIAN', 'CONTRACTOR TECHNICIAN', 'MAINTENANCE SUPERVISOR', 'CONTRACTOR SUPERVISOR')] <- 'MAINTENANCE'
df15_no_out$groupname_bin[df15_no_out$groupname %in% c('ACCOUNTING', 'PURCHASING MANAGER', 'SUPPLY CHAIN', 'STOREROOM MANAGER', 'STOREROOM CLERK', 'SENIOR STOREROOM CLERK')] <- 'SUPPLY CHAIN'

df15_no_out <- mutate_at(df15_no_out, vars('groupname_bin'), as.factor)

df15_no_out$gname_bin_num <- ifelse(df15_no_out$groupname_bin == "IT", 1,
                                    ifelse(df15_no_out$groupname_bin == "MAINTENANCE", 2,
                                           ifelse(df15_no_out$groupname_bin == "OPERATIONS", 3,
                                                  ifelse(df15_no_out$groupname_bin == "PIPELINE", 4,
                                                         ifelse(df15_no_out$groupname_bin == "PLANNING", 5,
                                                                ifelse(df15_no_out$groupname_bin == "SUPPLY CHAIN", 6,
                                                                       NA ))))))

df16_no_out <- df15_no_out
```

```{r }
#--------------- Initial data manipulation  ----------------------
#Start with dataframe: no outliers, and OK_all not present
#sessiontime_num - expresses sessiontime in numeric format

#Create new dataframe with select columns
df17 <- df16_no_out %>% select(maxsessionuid, defsite_grp, sessiontime_num)

#Select data where defsite_grp = TX_all
df17_TX <- df17 %>% filter(defsite_grp == "TX_all")

#Histogram of session time for TX_all
plot_hist = function(df, numcols, bins = 10){
  options(repr.plot.width=4, repr.plot.height=3) # Set the initial plot area dimensions
  for(col in numcols){
    if(is.numeric(df[,col])){
      bw = (max(df[,col]) - min(df[,col]))/(bins + 1)
      p = ggplot(df, aes_string(col)) + ggtitle("Sessiontime_num (mins) - TX") +
        geom_histogram(alpha = 0.6, binwidth = bw) 
      print(p)
    }
  }
}
numcols = c('sessiontime_num')
plot_hist(df17_TX, numcols)

#Select data where defsite_grp = LA_all
df17_LA <- df17 %>% filter(defsite_grp == "LA_all")

#Histogram of session time for LA_all
plot_hist = function(df, numcols, bins = 10){
  options(repr.plot.width=4, repr.plot.height=3) # Set the initial plot area dimensions
  for(col in numcols){
    if(is.numeric(df[,col])){
      bw = (max(df[,col]) - min(df[,col]))/(bins + 1)
      p = ggplot(df, aes_string(col)) + ggtitle("Sessiontime_num (mins) - LA") +
        geom_histogram(alpha = 0.6, binwidth = bw) 
      print(p)
    }
  }
}
numcols = c('sessiontime_num')
plot_hist(df17_LA, numcols)

# ------------ NULL HYPOTHESIS testing - sessiontime_num for LA and TX -------------------- 

#Convert table to dataframe - for LA_all
class(df17_LA)
df17_LA <- as.data.frame(df17_LA)

#Calculate mean of hourly LOGIN frequency - for LA_all
mean(df17_LA$sessiontime_num) 

#Convert table to dataframe - for TX_all
df17_TX <- as.data.frame(df17_TX)

#Calculate mean of hourly LOGIN frequency - for TX_all
mean(df17_TX$sessiontime_num) 

#Diff in mean of frequency of hourly LOGINs - TX vs LA
diff_mean_freq = round((mean(df17_TX$sessiontime_num) - mean(df17_LA$sessiontime_num)), 3)
diff_mean_freq

#Combine df17_LA and df17_TX
df18_TX_LA <- rbind(df17_LA, df17_TX)

table(df18_TX_LA$defsite_grp)

```
**Observation:**    
1. Have 13736 samples for LA and 7929 samples from TX.

```{r }
#Run a two-group t-test to determine if differences in means are significant
#default in R is to run the "Welch" version of the test.
#Welch - does not make any assumptions about the variances of the two groups
t.test(sessiontime_num ~ defsite_grp, data= df18_TX_LA)

```
**Observations:**  
1. # t-value = -18.6 - i.e.,the difference in means is ~18 times larger than would be expected by chance.  
2. How often would a result this big happen if the null hypothesis was actually true?  
3. Because p < 2.2e-16; i.e., atleast 2.2*10-14% of the time.  
4. This is sufficient to reject the NULL hypothesis - difference in means of session times between TX and LA users didn't happen by chance.

```{r }
#Display data on box plot
ggplot(df18_TX_LA, aes(x=defsite_grp, y=sessiontime_num, fill = defsite_grp)) + 
  geom_boxplot() +
  stat_summary(fun=mean, geom="point", shape=23, size=4)

```

```{r }
# Estimate the effect size (Cohen's d)
cohen.d(sessiontime_num ~ defsite_grp, data=df18_TX_LA)

```
**Observations:**  
1. tells us that our sample effect size is small (d estimate = -0.26 (small)).  
2. 95% confident that true effect size is somewhere between small and very small.  
